---
title: "Financial Econometrics: Equity Return Predictability & Interest Rate Dynamics"
author: "Dylan McClish"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This analysis applies econometric methods to two core problems in quantitative finance:

1. **Equity return predictability** — does the S&P 500 P/E ratio (and dividend yield) have
   predictive power over future returns at horizons from 1 month to 4 years?
2. **Interest rate dynamics** — can the Vasicek and CIR mean-reversion models be calibrated
   to 60+ years of U.S. Treasury yields, and how do exact MLE and Euler approximations compare?

---

# Part I — S&P 500 Return Predictability

## Load Data

```{r load-equity-data}
library(ggplot2)

sp500_monthly <- read.csv("sp_historical.csv")
sp500_pe      <- read.csv("sp500_pe_ratios_1871_2018.csv")
div_yield_df  <- read.csv("sp500_dividend_yield.csv")
```

## Compute Forward Returns

```{r compute-returns}
sp500_monthly$Date <- as.Date(sp500_monthly$Date, format = "%m/%d/%Y")
colnames(sp500_monthly)[2] <- "Close"

price <- sp500_monthly$Close

# h-month forward return at time t: price[t+h]/price[t] - 1
sp500_monthly$ret_1m  <- c(rep(NA, 1),  price[2:length(price)]  / price[1:(length(price)-1)]  - 1)
sp500_monthly$ret_3m  <- c(rep(NA, 3),  price[4:length(price)]  / price[1:(length(price)-3)]  - 1)
sp500_monthly$ret_12m <- c(rep(NA, 12), price[13:length(price)] / price[1:(length(price)-12)] - 1)
sp500_monthly$ret_24m <- c(rep(NA, 24), price[25:length(price)] / price[1:(length(price)-24)] - 1)
sp500_monthly$ret_36m <- c(rep(NA, 36), price[37:length(price)] / price[1:(length(price)-36)] - 1)
sp500_monthly$ret_48m <- c(rep(NA, 48), price[49:length(price)] / price[1:(length(price)-48)] - 1)

# restrict to Jan 1981 – Dec 2018 as specified
sp500_monthly <- sp500_monthly[sp500_monthly$Date >= as.Date("1981-01-01") &
                                sp500_monthly$Date <= as.Date("2018-12-31"), ]
```

## P/E Ratio: 1871–2018

The Shiller CAPE (cyclically adjusted P/E) is one of the most-watched long-run valuation
metrics. Elevated P/E ratios historically correspond with lower subsequent returns.

```{r pe-time-series}
sp500_pe$Date <- as.Date(sp500_pe$Date)

ggplot(sp500_pe, aes(x = Date, y = PE_ratio)) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  geom_hline(yintercept = mean(sp500_pe$PE_ratio, na.rm = TRUE),
             linetype = "dashed", color = "firebrick", linewidth = 0.5) +
  labs(title = "S&P 500 P/E Ratio (1871–2018)",
       subtitle = "Dashed line = historical mean",
       x = NULL, y = "P/E Ratio") +
  theme_minimal()
```

## Part (a): E/P Ratio as a Return Predictor

We regress forward returns at each horizon h on the earnings-to-price ratio (E/P = 1/PE).
The key question: does a higher earnings yield today predict higher returns in the future?

```{r ep-regressions}
sp500_pe$Date <- as.Date(sp500_pe$Date)
df <- merge(sp500_monthly, sp500_pe[, c("Date", "PE_ratio")], by = "Date", all.x = TRUE)
df$EP <- 1 / df$PE_ratio

n        <- nrow(df)
horizons <- c(1, 3, 12, 24, 36, 48)
ret_cols <- c("ret_1m", "ret_3m", "ret_12m", "ret_24m", "ret_36m", "ret_48m")

ep_results <- data.frame(horizon = integer(), beta = numeric(),
                         t_stat = numeric(), r_squared = numeric())

for (i in seq_along(horizons)) {
  h       <- horizons[i]
  fwd_ret <- c(df[[ret_cols[i]]][(h + 1):n], rep(NA, h))
  fit     <- lm(fwd_ret ~ df$EP)
  s       <- summary(fit)
  ep_results <- rbind(ep_results, data.frame(
    horizon   = h,
    beta      = round(coef(s)[2, 1], 4),
    t_stat    = round(coef(s)[2, 3], 3),
    r_squared = round(s$r.squared, 4)
  ))
  cat(sprintf("\n--- %2d-Month Forward Return ~ E/P ---\n", h))
  print(s$coefficients)
  cat(sprintf("R²: %.4f\n", s$r.squared))
}
```

R² rises substantially with horizon — consistent with long-run mean reversion in equity
valuations. The 48-month regression picks up the most predictability.

## Part (b): Adding the Log Dividend-to-Price Ratio

We add log(D/P) as a second regressor. Together, E/P and log(D/P) capture complementary
dimensions of valuation: earnings power vs. cash returned to shareholders.

```{r dp-regressions}
div_yield_df$Date <- as.Date(div_yield_df$Date)
div_yield_df$ym   <- format(div_yield_df$Date, "%Y-%m")
df$ym             <- format(df$Date, "%Y-%m")
df <- merge(df, div_yield_df[, c("ym", "div_yield")], by = "ym", all.x = TRUE)
df <- df[order(df$Date), ]
df$log_dp <- log(df$div_yield)

n <- nrow(df)

for (i in seq_along(horizons)) {
  h       <- horizons[i]
  fwd_ret <- c(df[[ret_cols[i]]][(h + 1):n], rep(NA, h))
  fit     <- lm(fwd_ret ~ df$EP + df$log_dp)
  s       <- summary(fit)
  cat(sprintf("\n--- %2d-Month Forward Return ~ E/P + log(D/P) ---\n", h))
  print(s$coefficients)
  cat(sprintf("R²: %.4f\n", s$r.squared))
}
```

---

# Part II — Interest Rate Model Calibration

We fit three models to monthly 5-year U.S. Treasury constant-maturity yields from FRED,
spanning January 1962 – December 2023 (~740 monthly observations). The Vasicek and CIR
models are both mean-reverting diffusion processes; the key difference is that CIR's volatility
scales with √r, preventing negative rates.

## Load Treasury Yield Data

```{r load-rates}
gs5_raw <- read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?id=GS5")
colnames(gs5_raw) <- c("DATE", "GS5")
gs5_raw$DATE <- as.Date(gs5_raw$DATE)
gs5_raw$GS5  <- suppressWarnings(as.numeric(gs5_raw$GS5))

gs5_raw <- gs5_raw[gs5_raw$DATE >= as.Date("1962-01-01") &
                   gs5_raw$DATE <= as.Date("2023-12-31"), ]
gs5_raw <- na.omit(gs5_raw)

ggplot(gs5_raw, aes(x = DATE, y = GS5)) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  labs(title = "5-Year U.S. Treasury Yield (1962–2023)",
       x = NULL, y = "Yield (%)") +
  theme_minimal()

r     <- gs5_raw$GS5 / 100
delta <- 1 / 12
n     <- length(r)
r_t   <- r[-n]
r_t1  <- r[-1]
```

## Model 1 — Vasicek Exact MLE

The Vasicek model is:
  dr_t = κ(α − r_t) dt + σ dW_t

The transition density r_{t+δ} | r_t is Gaussian, so exact MLE reduces to OLS of r_{t+1}
on r_t. We back out κ, α, σ from the OLS coefficients via the analytical mapping.

```{r vasicek-mle}
fit_a   <- lm(r_t1 ~ r_t)
d_hat   <- coef(fit_a)[["r_t"]]
c_hat   <- coef(fit_a)[["(Intercept)"]]
s2      <- mean(residuals(fit_a)^2)

kappa_a <- -log(d_hat) / delta
alpha_a <- c_hat / (1 - d_hat)
sigma_a <- sqrt(s2 * 2 * kappa_a / (1 - exp(-2 * kappa_a * delta)))

cat("Vasicek Exact MLE\n")
cat(sprintf("  κ (mean-reversion speed): %.4f\n", kappa_a))
cat(sprintf("  α (long-run mean):        %.4f  (%.2f%%)\n", alpha_a, alpha_a * 100))
cat(sprintf("  σ (diffusion):            %.4f\n", sigma_a))
```

## Model 2 — Vasicek Euler Approximation

The Euler linearization of the Vasicek SDE gives:
  Δr_t ≈ κα·δ − κδ·r_t + σ√δ·ε_t

We regress Δr on r_t and recover parameters from the intercept and slope.

```{r vasicek-euler}
delta_r <- diff(r)
fit_b   <- lm(delta_r ~ r_t)
A_b     <- coef(fit_b)[["(Intercept)"]]
B_b     <- coef(fit_b)[["r_t"]]

kappa_b <- -B_b / delta
alpha_b <- -A_b / B_b
sigma_b <- sd(residuals(fit_b)) / sqrt(delta)

cat("Vasicek Euler Approximation\n")
cat(sprintf("  κ: %.4f\n", kappa_b))
cat(sprintf("  α: %.4f  (%.2f%%)\n", alpha_b, alpha_b * 100))
cat(sprintf("  σ: %.4f\n", sigma_b))
```

## Model 3 — CIR Euler Approximation

The CIR model is:
  dr_t = κ(α − r_t) dt + σ√r_t dW_t

Dividing the Euler equation by √r_t homogenizes the variance, yielding a no-intercept
regression with regressors 1/√r_t and √r_t.

```{r cir-euler}
y_c   <- delta_r / sqrt(r_t)
z1    <- 1 / sqrt(r_t)
z2    <- sqrt(r_t)
fit_c <- lm(y_c ~ z1 + z2 - 1)

A_c     <- coef(fit_c)[["z1"]]
B_c     <- coef(fit_c)[["z2"]]
kappa_c <- -B_c / delta
alpha_c <- -A_c / B_c
sigma_c <- sd(residuals(fit_c)) / sqrt(delta)

cat("CIR Euler Approximation\n")
cat(sprintf("  κ: %.4f\n", kappa_c))
cat(sprintf("  α: %.4f  (%.2f%%)\n", alpha_c, alpha_c * 100))
cat(sprintf("  σ: %.4f\n", sigma_c))
```

## Model Comparison

```{r comparison-table}
comparison <- data.frame(
  Model  = c("Vasicek (Exact MLE)", "Vasicek (Euler)", "CIR (Euler)"),
  kappa  = round(c(kappa_a, kappa_b, kappa_c), 4),
  alpha  = round(c(alpha_a, alpha_b, alpha_c) * 100, 3),
  sigma  = round(c(sigma_a, sigma_b, sigma_c), 4)
)
colnames(comparison) <- c("Model", "κ", "α (long-run mean, %)", "σ")
knitr::kable(comparison, align = "lrrr")
```

All three models estimate a long-run mean near the historical average Treasury yield.
The Vasicek exact MLE and Euler estimates are nearly identical (as expected for monthly data
with small δ), while the CIR estimate reflects its different noise structure. The speed of
mean reversion κ implies a half-life of roughly log(2)/κ months.
