---
title: "Stylized Facts in Financial Time Series: Empirical Properties, Autocorrelation Tests, and Nonlinear Dependence"
author: "Dylan McClish"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5)
```

This analysis examines the statistical properties of equity returns in three parts:

1. **Empirical stylized facts** — distributional and autocorrelation properties of NASDAQ and
   IBM returns across daily, weekly, and monthly frequencies
2. **Validation of white-noise tests** — Monte Carlo simulation confirming that the
   Portmanteau (Box-Pierce) and Ljung-Box Q statistics converge to their theoretical
   chi-squared distributions as sample size grows
3. **Nonlinear dependence** — a time series with zero linear autocorrelation but strong
   nonlinear structure, illustrating the limits of ACF-based tests

---

# Part I — Empirical Properties of NASDAQ and IBM Returns

## Load and Process Data

```{r load-data}
library(ggplot2)

# helper: load + clean a price CSV, compute simple and log returns
load_returns <- function(file, date_fmt = "%b %d, %Y") {
  df <- read.csv(file, stringsAsFactors = FALSE)
  df$price <- as.numeric(gsub("[$,]", "", trimws(df$Close)))
  df$Date  <- as.Date(df$Date, format = date_fmt)
  df <- df[order(df$Date), ]
  df$simple_ret <- c(NA, diff(df$price) / head(df$price, -1))
  df$log_ret    <- c(NA, diff(log(df$price)))
  df
}

nasdaq_daily   <- load_returns("nasdaq_daily.csv")
nasdaq_weekly  <- load_returns("nasdaq_weekly.csv")
nasdaq_monthly <- load_returns("nasdaq_monthly.csv")

ibm_daily      <- load_returns("ibm_daily.csv")
ibm_weekly     <- load_returns("ibm_weekly.csv")
ibm_monthly    <- load_returns("ibm_monthly.csv")
```

---

## 1A — Price Levels and Log Return Time Series

Plotting price levels against log returns illustrates the signature volatility clustering of
financial markets: large moves tend to cluster together in time regardless of sign.

```{r ts-plots}
par(mfrow = c(1, 2))
plot(nasdaq_daily$Date, nasdaq_daily$price, type = "l", col = "steelblue",
     main = "NASDAQ Daily Prices", xlab = "", ylab = "Index Level")
plot(nasdaq_daily$Date[-1], na.omit(nasdaq_daily$log_ret), type = "l", col = "steelblue",
     main = "NASDAQ Daily Log Returns", xlab = "", ylab = "Log Return")
par(mfrow = c(1, 1))
```

```{r ibm-ts}
par(mfrow = c(1, 2))
plot(ibm_daily$Date, ibm_daily$price, type = "l", col = "darkorange",
     main = "IBM Daily Prices", xlab = "", ylab = "Price ($)")
plot(ibm_daily$Date[-1], na.omit(ibm_daily$log_ret), type = "l", col = "darkorange",
     main = "IBM Daily Log Returns", xlab = "", ylab = "Log Return")
par(mfrow = c(1, 1))
```

---

## 1B — Simple vs. Log Returns

Log returns (r_t = log P_t/P_{t-1}) and simple returns (R_t = P_t/P_{t-1} − 1) are nearly
identical for small moves but diverge for large ones. The tight 45° relationship confirms the
log approximation r ≈ R holds well for daily and weekly data; monthly data shows more spread.

```{r simple-vs-log}
plot_simple_vs_log <- function(df, title) {
  r <- na.omit(data.frame(simple = df$simple_ret, log = df$log_ret))
  plot(r$simple, r$log, pch = 20, cex = 0.4, col = rgb(0,0,0,0.3),
       xlab = "Simple Return", ylab = "Log Return", main = title)
  abline(0, 1, col = "red", lty = 2)
}

par(mfrow = c(2, 3))
plot_simple_vs_log(nasdaq_daily,   "NASDAQ Daily")
plot_simple_vs_log(nasdaq_weekly,  "NASDAQ Weekly")
plot_simple_vs_log(nasdaq_monthly, "NASDAQ Monthly")
plot_simple_vs_log(ibm_daily,      "IBM Daily")
plot_simple_vs_log(ibm_weekly,     "IBM Weekly")
plot_simple_vs_log(ibm_monthly,    "IBM Monthly")
par(mfrow = c(1, 1))
```

---

## 1C — Autocorrelation: Returns, Squared Returns, Absolute Returns

A key stylized fact: **log returns are nearly uncorrelated** (consistent with weak-form
efficiency), but **squared and absolute returns exhibit significant positive autocorrelation**
(volatility clustering). This contrast is the empirical motivation for GARCH-type models.

```{r acf-plots}
plot_acf_triple <- function(df, label) {
  r  <- na.omit(df$log_ret)
  r2 <- r^2
  ar <- abs(r)
  par(mfrow = c(1, 3))
  acf(r,  lag.max = 30, main = paste(label, "— Log Returns"))
  acf(r2, lag.max = 30, main = paste(label, "— Squared Returns"))
  acf(ar, lag.max = 30, main = paste(label, "— Absolute Returns"))
  par(mfrow = c(1, 1))
}

plot_acf_triple(nasdaq_daily,   "NASDAQ Daily")
plot_acf_triple(nasdaq_weekly,  "NASDAQ Weekly")
plot_acf_triple(nasdaq_monthly, "NASDAQ Monthly")
plot_acf_triple(ibm_daily,      "IBM Daily")
```

The near-zero ACF of raw returns contrasts sharply with the persistent ACF of squared and
absolute returns — the latter two are proxies for volatility, and their correlation structure
is the empirical fingerprint of ARCH effects.

---

## 1D — Distributional Properties: Histograms and Q-Q Plots

```{r dist-plots}
plot_dist <- function(df, label) {
  r <- na.omit(df$log_ret)
  par(mfrow = c(1, 2))
  hist(r, breaks = 50, freq = FALSE, col = "lightblue", border = "white",
       main = paste(label, "— Distribution"), xlab = "Log Return")
  curve(dnorm(x, mean(r), sd(r)), add = TRUE, col = "red", lwd = 2)
  qqnorm(r, main = paste(label, "— Q-Q Plot"), pch = 20, cex = 0.5)
  qqline(r, col = "red")
  par(mfrow = c(1, 1))
}

plot_dist(nasdaq_daily,   "NASDAQ Daily")
plot_dist(nasdaq_weekly,  "NASDAQ Weekly")
plot_dist(nasdaq_monthly, "NASDAQ Monthly")
plot_dist(ibm_daily,      "IBM Daily")
```

The S-shaped Q-Q plot deviation at both tails confirms **leptokurtosis** (fat tails) at
all frequencies — daily returns deviate most severely, consistent with the finding that
the normal approximation improves as frequency decreases (CLT aggregation effect).

---

# Part II — Monte Carlo Validation of Autocorrelation Tests

Before applying white-noise tests to real data, we verify that the test statistics converge
to their theoretical null distributions. We simulate 400 i.i.d. N(0,1) series and check:

- **Z = √T · ρ̂(1)**: should converge to N(0,1)
- **Q_m (Box-Pierce)**: should converge to χ²(m)
- **Q*_m (Ljung-Box)**: a finite-sample corrected version, also targeting χ²(m)

at lag lengths m ∈ {3, 6, 12} and sample sizes T ∈ {100, 500}.

```{r simulation-setup}
set.seed(42)
n_sim    <- 400
m_values <- c(3, 6, 12)

run_simulation <- function(T) {
  Z     <- numeric(n_sim)
  Q     <- lapply(seq_along(m_values), function(i) numeric(n_sim))
  Q_star <- lapply(seq_along(m_values), function(i) numeric(n_sim))

  for (sim in 1:n_sim) {
    x       <- rnorm(T)
    acf_all <- acf(x, lag.max = 12, plot = FALSE)$acf[-1]
    rho1    <- acf_all[1]
    Z[sim]  <- sqrt(T) * rho1

    for (j in seq_along(m_values)) {
      m          <- m_values[j]
      Q[[j]][sim]      <- T * sum(acf_all[1:m]^2)
      Q_star[[j]][sim] <- T * (T + 2) * sum(acf_all[1:m]^2 / (T - 1:m))
    }
  }
  list(Z = Z, Q = Q, Q_star = Q_star)
}

sim100 <- run_simulation(100)
sim500 <- run_simulation(500)
```

## Convergence of Z Statistic to N(0,1)

```{r z-histograms}
par(mfrow = c(1, 2))
for (res in list(list(sim100$Z, "T = 100"), list(sim500$Z, "T = 500"))) {
  hist(res[[1]], breaks = 30, freq = FALSE, col = "lightblue", border = "white",
       main = paste("Z =", expression(sqrt(T)), "· ρ̂(1),", res[[2]]),
       xlab = "Z")
  curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
}
par(mfrow = c(1, 1))
```

## Convergence of Q_m and Q*_m to χ²(m)

```{r q-histograms}
plot_q_hist <- function(res, T_label) {
  par(mfrow = c(2, 3))
  for (j in seq_along(m_values)) {
    m <- m_values[j]
    hist(res$Q[[j]], breaks = 30, freq = FALSE, col = "lightgreen", border = "white",
         main = paste0("Q_", m, " (Box-Pierce), ", T_label), xlab = "Statistic")
    curve(dchisq(x, m), add = TRUE, col = "red", lwd = 2)

    hist(res$Q_star[[j]], breaks = 30, freq = FALSE, col = "lightyellow", border = "white",
         main = paste0("Q*_", m, " (Ljung-Box), ", T_label), xlab = "Statistic")
    curve(dchisq(x, m), add = TRUE, col = "red", lwd = 2)
  }
  par(mfrow = c(1, 1))
}

plot_q_hist(sim100, "T=100")
plot_q_hist(sim500, "T=500")
```

## Empirical vs. Theoretical Percentiles

```{r percentile-table}
cat("=== T = 100 ===\n")
cat(sprintf("Z:  90th=%.3f (theory %.3f)  95th=%.3f (%.3f)  99th=%.3f (%.3f)\n",
  quantile(sim100$Z, 0.90), qnorm(0.90),
  quantile(sim100$Z, 0.95), qnorm(0.95),
  quantile(sim100$Z, 0.99), qnorm(0.99)))

for (j in seq_along(m_values)) {
  m <- m_values[j]
  cat(sprintf("Q_%d: 95th=%.3f (theory %.3f)  99th=%.3f (%.3f)\n", m,
    quantile(sim100$Q[[j]], 0.95), qchisq(0.95, m),
    quantile(sim100$Q[[j]], 0.99), qchisq(0.99, m)))
  cat(sprintf("Q*_%d: 95th=%.3f (theory %.3f)  99th=%.3f (%.3f)\n", m,
    quantile(sim100$Q_star[[j]], 0.95), qchisq(0.95, m),
    quantile(sim100$Q_star[[j]], 0.99), qchisq(0.99, m)))
}

cat("\n=== T = 500 ===\n")
cat(sprintf("Z:  90th=%.3f (theory %.3f)  95th=%.3f (%.3f)  99th=%.3f (%.3f)\n",
  quantile(sim500$Z, 0.90), qnorm(0.90),
  quantile(sim500$Z, 0.95), qnorm(0.95),
  quantile(sim500$Z, 0.99), qnorm(0.99)))

for (j in seq_along(m_values)) {
  m <- m_values[j]
  cat(sprintf("Q_%d: 95th=%.3f (theory %.3f)  99th=%.3f (%.3f)\n", m,
    quantile(sim500$Q[[j]], 0.95), qchisq(0.95, m),
    quantile(sim500$Q[[j]], 0.99), qchisq(0.99, m)))
}
```

At T=100 the Ljung-Box Q* tracks the chi-squared critical values more accurately than the
plain Box-Pierce Q, confirming its finite-sample advantage. By T=500 both statistics are
essentially indistinguishable from the theoretical distribution.

---

# Part III — Nonlinear Dependence

A critical limitation of ACF-based tests: they detect **linear** autocorrelation only.
We simulate a nonlinear process:

  X_t = 0.8 · ε²_{t−1} / (1 + ε²_{t−1}) + ε_t

where ε_t is i.i.d. N(0,1). This process has **zero linear autocorrelation** (ACF will
look flat) but the conditional mean depends nonlinearly on the past shock magnitude —
structure invisible to Box-Pierce and Ljung-Box tests.

```{r nonlinear-series}
set.seed(42)
T       <- 400
epsilon <- rnorm(T)
X       <- numeric(T)
X[1]    <- epsilon[1]

for (t in 2:T)
  X[t] <- 0.8 * epsilon[t-1]^2 / (1 + epsilon[t-1]^2) + epsilon[t]

par(mfrow = c(1, 1))
plot(X, type = "l", col = "steelblue",
     main = expression("Nonlinear Series: " * X[t] == frac(0.8 * epsilon[t-1]^2, 1 + epsilon[t-1]^2) + epsilon[t]),
     xlab = "t", ylab = expression(X[t]))
```

```{r nonlinear-acf}
par(mfrow = c(1, 2))
acf(X,    lag.max = 30, main = "ACF of X_t  (appears white noise)")
acf(X^2,  lag.max = 30, main = "ACF of X²_t  (reveals nonlinear structure)")
par(mfrow = c(1, 1))
```

The ACF of X_t is indistinguishable from white noise — a standard Ljung-Box test would
fail to reject the null. But the ACF of X²_t exposes the hidden structure: the squared
series has significant positive autocorrelation, exactly analogous to the ARCH signature
seen in real return data in Part I. This illustrates why testing for ARCH effects requires
looking beyond first-moment autocorrelation.
